---
title: "Statistical Learning"
author: "Xinyu Yang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(viridis)
set.seed(11)
```

## LASSO
LASSO aims for good prediction accuracy, not necessarily interpretation/inference

```{r}
#bwt dataset
bwt_df = 
  read_csv("./data/birthweight.csv") %>% 
  janitor::clean_names() %>%
  mutate(
    babysex = as.factor(babysex),
    babysex = fct_recode(babysex, "male" = "1", "female" = "2"),
    frace = as.factor(frace),
    frace = fct_recode(frace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4", "other" = "8"),
    malform = as.logical(malform),
    mrace = as.factor(mrace),
    mrace = fct_recode(mrace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4")) %>% 
  sample_n(200)
```

Fit lasso model
```{r}
x = model.matrix(bwt ~ ., bwt_df)[,-1] #turn the dataframe into a matrix
y = bwt_df$bwt

lambda = 10^(seq(3, -2, -0.1))

lasso_fit =
  glmnet(x, y, lambda = lambda)

lasso_cv =
  cv.glmnet(x, y, lambda = lambda)

lambda_opt = lasso_cv$lambda.min
```


Plot the output:
```{r}
broom::tidy(lasso_fit) %>% 
  select(term, lambda, estimate) %>% 
  complete(term, lambda, fill = list(estimate = 0) ) %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate, group = term, color = term)) + 
  geom_path() + 
  geom_vline(xintercept = log(lambda_opt, 10), color = "blue", size = 1.2) +
  theme(legend.position = "none")
```

The plot of lamda and estimate of parameters:
When lamda is small, everything was included in the model; as lamda increases, the estimate of paramters decrease (closer and closer to 0) and eventually all equal to 0 when lamda reaches its maximum.

Blue line -> smallest prediction error (correponding to the lowest point in the cross validation curve, see below)

Cross-validation:
```{r}
broom::tidy(lasso_cv) %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate)) + 
  geom_point() 
```


## CLustering K-Means

Clustering partitions your dataset 

```{r}
#pokemon dataset
poke_df = 
  read_csv("./data/pokemon.csv") %>% 
  janitor::clean_names() %>% 
  select(hp, speed)

poke_df %>% 
  ggplot(aes(x = hp, y = speed)) + 
  geom_point()
```

The old base-R k-means function
```{r}
kmeans_fit =
  kmeans(x = poke_df, centers = 3)
```

Use the broom package to tidy it..
```{r}
poke_df =
  broom::augment(kmeans_fit, poke_df)

poke_df %>% 
  ggplot(aes(x = hp, y = speed, color = .cluster)) +
  geom_point()
```


Re-fit the model using different numbers of clusters (1~4)
```{r}
clusts =
  tibble(k = 2:4) %>%
  mutate(
    km_fit =    map(k, ~kmeans(poke_df, .x)),
    augmented = map(km_fit, ~broom::augment(.x, poke_df))
  )

clusts %>% 
  select(-km_fit) %>% 
  unnest(augmented) %>% 
  ggplot(aes(hp, speed, color = .cluster)) +
  geom_point(aes(color = .cluster)) +
  facet_grid(~k)
```

## CLustering - trajectories
SAS: proc traj

```{r}
traj_data = 
  read_csv("./data/trajectories.csv")


traj_data %>% 
  ggplot(aes(x = week, y = value, group = subj)) + 
  geom_point() + 
  geom_path()
```


Create the slope and intercept for each individual across week (value ~ week)
```{r}
int_slope_df = 
  traj_data %>% 
  nest(data = week:value) %>% 
  mutate(
    models = map(data, ~lm(value ~ week, data = .x)),
    result = map(models, broom::tidy)
  ) %>% 
  select(subj, result) %>% 
  unnest(result) %>% 
  select(subj, term, estimate) %>% 
  pivot_wider(
    names_from = term,
    values_from = estimate
  ) %>% 
  rename(int = "(Intercept)", slope = week)

int_slope_df %>% 
  ggplot(aes(x = int, y = slope)) + 
  geom_point()
```

From the scatterplot it seems that there are two groups going on; we can use K-menas to extract that pattern

But first we need to remove the scaling of the two parameters (i.e. slope and intercept)
```{r}
km_fit = 
  kmeans(
    x = int_slope_df %>% select(-subj) %>% scale, 
    centers = 2)
```

```{r}
int_slope_df =
  broom::augment(km_fit, int_slope_df)

int_slope_df %>% 
  ggplot(aes(x = int, y = slope, color = .cluster)) +
  geom_point()
```

Re-do the first speghetti plot now grouped by the clusters
```{r}
left_join(traj_data, int_slope_df) %>% 
  ggplot(aes(x = week, y = value, group = subj, color = .cluster)) + 
  geom_point() + 
  geom_path() 
```

